The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-06 write-up/covid-writeup.tex
styles needed

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{soul,color,hyperref}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage[margin=1.5in]{geometry}

\numberwithin{equation}{section}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{example}{Example}
\newtheorem{remark}{Remark}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}

@@

\subsection{Regrettable rates}
\label{section:rates}

@@

Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.  This means selection bias and measurement error can have paradoxical effects.  Here we consider the ratio estimators for the relative change in the prevalence rate.  We let $\bar Y_1$ and $\bar Y_2$ denotes the prevalence at time step $1$ and $2$ respectively (i.e., prevalence on two consecutive days).  Then the ratio estimator is given by $r = \bar y_1 / \bar y_2$.  We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error. If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_1/\bar y_2 = y_1/y_2$).  We wish to express the error $\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}$ in statistical terms.  Using a Taylor series approximation, we show that the error can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_2}{\bar Y_1} &\times \bigg[ \rho_{I_2,Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)  -\rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \bigg] \\
&\times \left[ 1 - \rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.  The magnitude of the error depends on the population-level ratio $\bar Y_2 / \bar Y_1$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate. The second term represents the ``cancellation'' that the naive analyst is hoping will occur.  The analyst's intuition only holds when the difference is equal to zero.  This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all held constant.  In the case of New York, for example, recent weeks it seems plausible that $\Delta$ (i.e., selection bias) has not changed drastically.  Figure XX shows the rate of testing remains constant and the total cases relative to the population is small enough such that $f_1 \equiv f_2$ is reasonable. Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported (as far as this author knows).  We will show how such a setting leads to over-estimation prior to and under-estimation after peak prevalence.  The final component is an adjustment to this differential error based on the statistical error at the first time point.

@@

\subsection{Cross-country comparisons}

So far we have focused on understanding the disease in a single location and the limitations of using case-count data to understand population quantities of interest.  Many in the

So we can't estimate prevalence.  Can we not get a decent comparison?
$$
y_1 - y_2 = \frac{N_1}{n_1} ( \bar y_1 - Y_{1}) + \left( \frac{N_1}{n_1} Y_{1} - \frac{N_2}{n_2} Y_2  \right).
$$
OK then we should compare relative counts?
The question is then ``What are you comparing at a populatoin level?''
$$
y_1/N_1 - y_2/ N_2 = \left[ f_1 \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}} D_{M_1} - f_2 \sigma_{Y_2} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}}  \right] + \left[ f_1 \bar Y_1 - f_2 \bar Y_2 \right]
$$
Is this a quantity of genuine interest? Under SRS, this makes sense if $f_1 = f_2$.  Otherwise, you are comparing apples and oranges.  With selection bias, it's not clear that agreement in this quantity is even desirable.

@@

>>>>

 \documentclass[11pt]{amsart}

\RequirePackage[OT1]{fontenc}

\usepackage{biblatex}
%\bibliographystyle{apalike}

\usepackage{tchdr}
\boldshortcuts

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{soul,color,hyperref}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage[margin=1.5in]{geometry}

% \numberwithin{equation}{section}
% \theoremstyle{plain}
% \newtheorem{theorem}{Theorem}[section]
% \newtheorem{example}{Example}
% \newtheorem{remark}{Remark}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{criterion}[theorem]{Criterion}
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{exercise}[theorem]{Exercise}
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{proposition}[theorem]{Proposition}

@@

\subsection{Regrettable rates: complex biases resulting from self-selection}
\label{section:rates}

@@

Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.  This means selection bias and measurement error can have paradoxical effects.  Here we consider the ratio estimators for the relative change in the prevalence rate.  We let $\bar Y_{t-1}$ and $\bar Y_{t}$ denotes the prevalence at time step $t-1$ and $t$ respectively (i.e., prevalence on two consecutive days).  Then the ratio estimator is given by $r = \bar y_t / \bar y_{t-1}$.  We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error. If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_1/\bar y_2 = y_1/y_2$).  We wish to express the error $\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}$ in statistical terms.  Using a Taylor series approximation, we show that the error can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_2}{\bar Y_1} &\times \bigg[ \rho_{I_2,Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV (Y_2)  -\rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \bigg] \\
&\times \left[ 1 - \rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.  The magnitude of the error depends on the population-level ratio $\bar Y_2 / \bar Y_1$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate. The second term represents the ``cancellation'' that the naive analyst is hoping will occur.  The analyst's intuition only holds when the difference is equal to zero.  This occurs at times when data quality, sampling fraction, measurement-error, and prevalence are all held constant.  In the case of New York, for example, recent weeks it seems plausible that $\Delta$ (i.e., selection bias) has not changed drastically.  Figure XX shows the rate of testing remains constant and the total cases relative to the population is small enough such that $f_1 \equiv f_2$ is reasonable. Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported (as far as this author knows).  We will show how such a setting leads to over-estimation prior to and under-estimation after peak prevalence.  The final component is an adjustment to this differential error based on the statistical error at the first time point.

@@

\section{Cross-population comparisons}

So far we have focused on understanding the limitations of using case-count data to understand population quantities of interest for a \emph{single} population (i.e., United States, Germany, France or Michigan, New York, London in isolation).  Many are interested in cross-population comparisons.  Many pieces in the media as well as academic articles have plots of case-count over time aligned by time of first known case.  Others have claimed such comparisons unfair due to unequal population sizes and instead plotted case-count per million.  In this section, we focus on the statistical issues relevant to such comparisons. We do not make any comments on the multi-causal nature of success/failure in countries.

\subsection{Prevalence comparisons}

At a fixed time, suppose prevalence estimates $\bar y_1$ and $\bar y_2$ are observed for two populations. We can express the error $\bar y_1 - \bar y_2$ in statistical terms as
$$
(\bar Y_1 - \bar Y_2) + \left[ \rho_{I_1, Y_1} \times \sqrt{\frac{1-f_1}{f_1}} \times \sigma_{Y_1}  \right]
$$

We could instead think about the number of new cases $K_1$ and $K_2$ within each population.  Then
$$
y_1 - y_2 = \frac{N_1}{n_1} ( \bar y_1 - Y_{1}) + \left( \frac{N_1}{n_1} Y_{1} - \frac{N_2}{n_2} Y_2  \right).
$$
OK then we should compare relative counts?
The question is then ``What are you comparing at a populatoin level?''
$$
y_1/N_1 - y_2/ N_2 = \left[ f_1 \sigma_{Y_1} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}} D_{M_1} - f_2 \sigma_{Y_2} \rho_{I_1, Y_1} \sqrt{\frac{1-f_1}{f_1}}  \right] + \left[ f_1 \bar Y_1 - f_2 \bar Y_2 \right]
$$
Is this a quantity of genuine interest? Under SRS, this makes sense if $f_1 = f_2$.  Otherwise, you are comparing apples and oranges.  With selection bias, it's not clear that agreement in this quantity is even desirable.

@@

\subsection{Rate comparisons}
